---
title: "Using fastcramer: comparing two multivariate samples at scale"
author: "Alexander C. Furnas"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Using fastcramer: comparing two multivariate samples at scale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
set.seed(20261002)
```

## The question

Suppose each observation is a vector, such as a text embedding of a document, and
observations belong to one of two groups. Do the two groups occupy different
regions of the space? The Cramér two-sample test (Baringhaus & Franz 2004)
answers this without a parametric model. Its statistic compares the mean
distance between the groups with the mean distances within each group,

$$T = \frac{mn}{m+n}\left[\frac{2}{mn}\sum_{i,j}\phi(|x_i-y_j|^2)
      - \frac{1}{m^2}\sum_{i,i'}\phi(|x_i-x_{i'}|^2)
      - \frac{1}{n^2}\sum_{j,j'}\phi(|y_j-y_{j'}|^2)\right],$$

with $\phi(z) = \sqrt{z}/2$ by default, so that $T$ is zero when the two
groups are interchangeable and grows with their separation.

`fastcramer` computes this statistic exactly as `cramer::cramer.test()` does,
but roughly a thousand times faster, which changes what is practical: whole
groups of thousands of observations instead of subsamples of a few hundred.

## Simulated embeddings

We simulate 768-dimensional embeddings for two groups whose means differ
slightly along a random direction, with a nuisance structure of "topics" that
shifts both groups together.

```{r simulate}
library(fastcramer)

d <- 768
n_topics <- 12
topic_centres <- matrix(rnorm(n_topics * d, sd = 0.5), n_topics)
group_axis <- rnorm(d); group_axis <- group_axis / sqrt(sum(group_axis^2))

simulate_group <- function(n, shift, topic_probs) {
  topic <- sample(n_topics, n, replace = TRUE, prob = topic_probs)
  emb   <- topic_centres[topic, ] + matrix(rnorm(n * d), n) + shift * matrix(group_axis, n, d, byrow = TRUE)
  list(emb = emb, topic = topic)
}

# group A: 3,000 observations, spread over topics; group B: 600, concentrated
# in a few topics AND shifted along the group axis
A <- simulate_group(3000, shift = 0,   topic_probs = rep(1, n_topics))
B <- simulate_group(600,  shift = 0.6, topic_probs = c(rep(3, 4), rep(1, n_topics - 4)))
```

## The statistic

```{r statistic}
cramer_statistic(A$emb, B$emb)
system.time(cramer_statistic(A$emb, B$emb))
```

All five kernels of the `cramer` package are available and agree with it:

```{r kernels}
sapply(c("cramer", "bahr", "log", "fraca", "fracb"),
       function(k) cramer_statistic(A$emb[1:300, ], B$emb[1:300, ], kernel = k))
```

```{r check-cramer, eval = requireNamespace("cramer", quietly = TRUE)}
cramer::cramer.test(A$emb[1:300, ], B$emb[1:300, ], just.statistic = TRUE)$statistic
```

## A permutation test

`cramer_test()` pools the samples and reassigns group labels at random. The
Monte Carlo p-value is the share of permutations with a statistic at least
as large as the observed one.

```{r test}
cramer_test(A$emb[1:400, ], B$emb[1:400, ], replicates = 199)
```

## Holding nuisance structure fixed with strata

The two groups also differ in topic composition. Is the separation only
topic sorting? Permuting labels *within topic* preserves each topic's group
composition, so the null distribution reflects what the statistic would be if
group membership carried no information beyond topic.

```{r blocked}
emb   <- rbind(A$emb, B$emb)
topic <- c(A$topic, B$topic)
res_free    <- cramer_test(A$emb[1:400, ], B$emb[1:400, ], replicates = 199)
res_blocked <- cramer_test(A$emb[1:400, ], B$emb[1:400, ], replicates = 199,
                           block = c(A$topic[1:400], B$topic[1:400]))
c(unblocked = res_free$p.value, within_topic = res_blocked$p.value)
```

The blocked null typically has larger statistics than the free null, because
topic composition is no longer "explained away" by the permutation, so the
test is more conservative. Here the shift along the group axis survives it.

## Very large samples: bootstrap draws against a blocked null

When the groups contain hundreds of thousands of observations, even a fast
statistic cannot be evaluated on the full data, and one would not want to:
with that many observations any difference is "significant". The design used
in the motivating application (Furnas, Gao, Yin and Wang 2026) instead compares
two distributions of replicates. Each *observed* replicate draws `draw` observations with
replacement from each group and computes the statistic; each *permuted*
replicate first shuffles the group labels within strata, then draws in the
same way. `cramer_bootstrap()` runs both and summarises them by

* the **exceedance probability**, the share of observed/permuted pairs in
  which the observed statistic is larger, and
* the **percentile interval** of the pairwise differences,

rather than by a test of the two replicate means, whose p-value would scale
with the number of replicates instead of with the data.

```{r bootstrap}
boot <- cramer_bootstrap(A$emb, B$emb, replicates = 100, draw = 500,
                         block = topic, seed = 1)
boot
```

```{r bootstrap-plot}
reps <- boot$replicates
hist(reps$statistic[reps$type == "permuted"], col = "grey80", border = NA,
     xlim = range(reps$statistic), main = "", xlab = "Cramér statistic")
hist(reps$statistic[reps$type == "observed"], col = adjustcolor("firebrick", 0.6),
     border = NA, add = TRUE)
legend("topright", fill = c("grey80", adjustcolor("firebrick", 0.6)),
       legend = c("within-topic permuted draws", "observed draws"), bty = "n")
```

The default `draw` is the smaller of 5,000 and the smaller group's size, so
small groups are not resampled far beyond what they contain. Note that the
statistic's scale grows with `draw` through the $mn/(m+n)$ factor: compare
observed and permuted draws of the same size, not raw statistics across
different designs.

## Performance notes

The kernel evaluates each block of the cross-product with BLAS and accumulates
the distances, kernel values and sums in one pass. It is bound by memory
bandwidth rather than by the BLAS, so a tuned BLAS helps little and several
parallel workers on one machine share bandwidth; expect roughly 0.3 s per
5,000-by-5,000 statistic alone and about 1.5 s per worker with ten workers.
For permutation designs over many groups, parallelise across groups with
`parallel::mclapply()` and keep one BLAS thread per worker.

## References

Baringhaus, L., & Franz, C. (2004). On a new multivariate two-sample test.
*Journal of Multivariate Analysis*, 88(1), 190–206.

Furnas, A. C., Gao, J., Yin, Y., & Wang, D. (2026). Partisan disparities in
the production and uptake of science. Working paper.
